### 2016-07-29 Projects/DelCo_2012/Report/DelCo_2012.tex
Began editing CAFS report

Full scale fire experiments designed to address those questions were conducted over a period of years from 2012 through 2015. The experimental arrangements and results are presented in this report. The results from both the gas cooling and the suppression of residential scale room fires showed the effectiveness of water and CAF hose streams to be similar. Under the test conditions both agents were effective. This study did not address the previously documented attributes of CAF for improved exposure protection and improved prevention of a fire re-kindle, relative to plain water.

@@

A number of fire departments throughout the US have acquired CAFS units for structural firefighting, but many have limited or stopped the use of these CAFS for structural firefighting due to operational challenges or due to lack of understanding related to fireground safety issues. This project is expected to produce a better understanding of the effectiveness and safety implications of CAF for structural firefighting through the various tasks associated with this project, including:

\begin{enumerate}
\item  Perform a literature review of available published information on the use of CAF for structural firefighting~\cite{Mitchell:1}
\item  Conduct a workshop that brings interested and experienced parties together to discuss CAF experience and to provide guidance on the proposed experimental research plan~\cite{Grant:2011} 
\item  Conduct and document well-instrumented large-scale enclosure fire tests using both CAF and plain water to evaluate the effectiveness of CAF for interior structural firefighting
\item  Conduct and document identified fire ground related experiments to evaluate differences between CAF and plain water hose lines related to issues such as effective hose stream throw and distribution, nozzle reaction force, the force to kink the hose, potential for slipping on a variety of surfaces and flow separation~\cite{Carracino:2013,Dicus:2013,LaPolla:2012}. 
\end{enumerate}

@@

The key measures were to understand the gas cooling and fire suppression capabilities of CAF compared to water. The experimental designs were developed based on a project planning workshop which included the discussion of incidents where firefighters had concerns or experienced minor burns while using CAF during interior attack operations in residential occupancies~\cite{Grant:2011}. Anecdotally, other fire departments or firefighters had similar incidents or impressions of CAF.

Compressed air foam (CAF) generated with a solution of water and a Class A foam concentrate has been in use by wildland fire agencies since the late 1970's. The wildland fire community found that the Class A foam reduced the surface tension of water, which improved water's ability to penetrate and soak into wildland fuels. A CAFS could turn a Class A foam solution into a blanket of bubbles that would allow the wildland firefighters to coat the fuels. The additional ``soak'' time provided by the foam layer allowed the water to penetrate the hot surface of a burning fuel, and also provided additional time for water to penetrate into fuels being pre-treated in an effort to prevent ignition. An added feature of the blanket of bubbles is that firefighters could see where they had applied the agent so that they could make a more effective application. In the wildland, water for firefighting is typically a precious resource, is in limited supply, and has to be carried to where it is needed. CAF increases the effectiveness of water, and may reduce the quantity required to suppress fires and prevent the ignition of adjacent fuels. Last but not least the compressed air adds energy to the water stream providing additional reach and volume to a hose stream with a relatively low mass flow rate of water. These CAF advantages combine to provide wildland firefighters with a tool that could be used to pre-treat homes and structures to protect them from an approaching fire, or could be used for an exterior (indirect) fire suppression attack on a structure.

In the decades that followed the development of CAFS, a number of urban/suburban fire departments across the country began to integrate CAFS into their apparatus. For example, the Los Angeles County Fire Department incorporated CAFS into their apparatus for wildland urban interface fires, but also examined using them for structural fire fighting. Other departments across the country began to experiment with CAFS systems for use in structural firefighting. In the few cases that were documented, the findings about the effectiveness of the CAFS were favorable, yet after decades of fire department trials, CAF has still not been widely incorporated into structural firefighting tactics~\cite{USFA:CAFS,Lorh:2002}. This is illustrated by the example of one community, which after putting a fleet of CAFS equipped fire engines into service, had issued an order that the CAFS should not be used after incidents occurred where firefighters were burned while performing interior suppression operations with CAF inside a structure.

There has never been a comprehensive scientific study on the use of CAF for structural fire fighting. Fire departments have conducted pilot programs to examine implementation and effectiveness of CAF, foam suppliers and equipment manufacturers have conducted minimally instrumented experiments, typically in acquired structures, and test laboratories and universities have examined small sections of the CAF puzzle, but none of these efforts has resulted in the body of knowledge needed by the fire service~\cite{Madrzykowski:3,USFA:CAFS,Colletti:1,Boyd:1,Tafreshi:1,Madrzykowski:4,Colletti:2,Tafreshi:2,Lorh:2002,Colletti:3,Kim:2012}.

Demonstration of Biodegradable, Environmentally Safe, Non-toxic Fire Suppression Liquids NISTIR 6191 was published in 1998~\cite{Madrzykowski:4}. That study examined a broad range of tests in order to determine those parameters that most critically effect fire-fighting performance.  The types of experiments conducted and the impact of additives relative to water are included in Table~\ref{tab:agent_tests}.

@@

Based on the results of the NIST 1998 study, the fire fighting agents, commonly referred to as Class A foams, were superior to water in reducing contact angle, cooling and penetrating wood based fuels, mass retention on the wood fuels and ignition inhibition of the wood fuels.  When applied in the form of foam, the water agent solutions provide enhanced radiation exposure protection to a structure relative to plain water. The foams enhanced the water by keeping the water in place on vertical surfaces such as walls and the reduced inter-facial surface tension allowed the water to penetrate absorbent fuels. These enhancements also make the agents valuable tools for mop-up. The other experiments conducted during this study: specific heat, drop size, wood crib suppression, smoke generation and Class D fire suppression yielded similar results for both Class A foams and plain water.

In 2005, The National Fallen Firefighters Foundation held the National Fire Service Research Agenda Symposium.  One of the needs identified was ``improved extinguishing agents and delivery systems.'' The report states that ``The primary objective of this research should be to reduce the exposure of firefighters to the risks of interior fire suppression by developing more effective extinguishing agents and systems.'' In wildland firefighting, CAFS is typically applied directly to the fuel from a remote location, typically a non-IDLH environment. In structural firefighting, crews typically  have to enter the building, which contains the fire, and move under or through hot combustion products and pre-heated areas of the building (IDLH environment) as they approach the fire. In many cases, before the firefighters are able to apply the suppression agent directly to the burning fuel, they may have to mitigate the hazard of the hot gas layer or try to cool interior surfaces as they move toward the fire. It is important that these distinctions and differences between exterior and interior firefighting with CAF are understood better.

Montgomery County (MD) Fire and Rescue Service (MCFRS) is one of the larger fire departments in the State of Maryland and was a committed user of CAF with 35 CAFS front line engines. MCFRS had experienced two incidents where firefighters received burns in residential structure fires whiles CAF was being used to fight the fire during an interior attack. Concerns were raised within the department regarding the ability of CAF to cool fire gases and absorb heat as well as plain water. As a result, on March 4, 2010, the Fire Chief of (MCFRS) issued a General Order prohibiting the use of CAF during offensive interior attacks. In this same order, the Chief directed the Division Chief of Operations to continue to evaluate the use of CAF during structure fires and to report back his findings and recommendations. That order was the catalyst for this study.

@@

All of the experiments were conducted at full scale in compartments and structures that were representative of room sizes in a residential occupancy. Technical panel members whose departments routinely use CAF for structural firefighting were invited and present at many of the fire suppression experiments. The series of experiments, planned during the workshop, for examining spray density measures, gas layer cooling, and fire suppression were conducted in the fall of 2012. The data from these experiments were presented to the technical panel member. The fire suppression experiments were conducted with a fuel package that included upholstered furniture. After a review of the data, a decision was made to conduct an additional series of suppression experiments with a fuel package that was composed mainly of wood based materials. The thinking was that the synthetic materials in the upholstered furniture burned only on the surface and were easily extinguished. Wood based fuels would char and present a deep seated fire challenge to the extinguishing agents.  After the wood based fuel package experiments were completed in 2013, it was decided in discussions with members of the technical panel that other experimental conditions needed to be examined: room size, additional ventilation, and a manually operated handline. The issue of a larger fire compartment had also been raised during the workshop discussion, but the desire for replicate experiments and financial limitations removed the large floor plan experiments from the original plan. To accomplish the last series of experiments, additional funding was obtained from NIST and the experiments were conducted in 2015. The fuel load contained both wood-based wall and ceiling finishes, pallets as well as carpeting and upholstered furniture.

The sections below describe the types of experiments, the experimental facilities, instrumentation, and fuel loads. Project partner, Montgomery County (MD) Fire and Rescue Service (MCFRS) provided the CAFS equipped engines and the staffing needed to provide both the water and CAF for the experiments. These experiments were conducted at the Delaware County Emergency Services Training Center (ESTC) in Sharon Hill, PA.

\section{Description of Experiments}
\label{sec:desc_experiments}
The experimental setup for this test series is described by the three types of experiments conducted: spray density measures, gas layer cooling, and fire suppression.  Baseline operating conditions were agreed upon at the CAFS workshop for these experiments (Table~\ref{tab:op_condition})~\cite{Grant:2011}. The nozzles chosen for these experiments were: Task Force Tips (TFT) Metro 1, fixed gallonage nozzles with a flow rate of 120~gpm. These are automatic nozzles which have a 7/8~inch slug-tip and can be varied from a straight stream to a wide fog for application. The set-up also included 100~ft of 1-3/4~inch hose. Nozzles were mounted to a ``blitz-fire'' monitor device and placed inside the structures. The nozzles were controlled and operated via a gate valve from the exterior. The exterior control had a gated wye to allow the flow of CAF through a hose directed away from the test structure. When the flow of CAF was needed for the experiment, the gated wye would be switched to allow the CAF to flow into the test structure. This was done to minimize, the build-up of``dry'' foam or a slug upon the initial flow of the nozzle into the test structure.

@@

The fire suppression series included 20 experiments in the single-story experimental structures and one two-story structure. Twelve experiments were conducted in the single story structures (6 in each structure) using both water and CAF application as a suppression agent. The nozzle patterns were varied and included; straight stream, solid stream, and narrow fog.  The two-story tests used both water and CAF from a straight-stream nozzle.

@@

The series of tests described within this report were conducted at the Delaware County Emergency Services Training Center, located in Sharon Hill, PA. A burn building as well as two purpose built concrete structures are located on the grounds of the ESTC. The burn building was used for the gas cooling experiments and the purpose built concrete structures were used for the fire suppression experiments.

@@

The burn building is a live fire training structure with both a two story and three story section. The building is supported with reinforced concrete beams and columns. The floors and ceilings are also concrete with the interior walls constructed of cement block. The walls and ceiling of the burn rooms are protected with a 25~mm thick layer of calcium silicate insulation and is covered by a 50~mm thick concrete tile. The floor of the burn rooms is protected with fire brick. Tests were conducted in the two-story, two-room configuration which has overall dimensions of 6.5~m (21.3~ft) by 9.2~m (30.2~ft) (cf. Fig.~\ref{fig:Delaware_County,_PA_Burn_Building_Layout}).

@@

\subsection{Concrete Block Structures}
\label{sec:Experimental Structures}
\subsubsection*{Single Story Structures}

@@

There are different components of uncertainty in the length, mass, temperature, heat flux, gas concentration, differential pressure, gas velocity and heat release rate reported here. Uncertainties are grouped into two categories according to the method used to estimate them. Type A uncertainties are those which are evaluated by statistical methods, and Type B are those which are evaluated by other means~\cite{Taylor&Kuyatt:1994}. Type B analysis of systematic uncertainties involves estimating the upper (+ a) and lower (- a) limits for the quantity in question such that the probability that the value would be in the interval ($\pm$ a) is essentially 100~\%. After estimating uncertainties by either Type A or B analysis, the uncertainties are combined in quadrature to yield the combined standard uncertainty. Then the combined standard uncertainty is multiplied by a coverage factor of two, which results in the expanded uncertainty with a 95~\% confidence interval (2$\sigma$).  For some of these components, such as the zero and calibration elements, uncertainties are derived from referenced instrument specifications. For other components, referenced research results and past experience with the instruments provided input in the uncertainty determination.

@@

There were 12 suppression tests conducted in the single story purpose-built concrete structure. Eight of the experiments used a fuel load comprised of typical household furnishings that included wooden wall paneling, two sofas, two overstuffed chairs, carpet, and padding (Section~\ref{sec:fire_suppression_furniture_fuel}). The other four experiments used wood pallet fuel load in a space with the ceiling and walls composed of wood based materials (Section~\ref{sec:fire_suppression_pallet_fuel}). Ignition occurred in the rear of the single story structures and the available ventilation was through the front doorway and for eight of the experiments a rear window was ``opened''to provide additional ventilation (see Table~\ref{tab:Test_Descriptions}). The fires were intended to replicate post-flashover residential room and contents fires. Hose lines were pre-deployed with fixed nozzles set in two locations: one in the hallway on the approach to the burn room (hallway nozzle) and one just inside the burn room to the right side of the hallway (room nozzle). Each nozzle flowed 15~s of water or compressed air foam depending on the test. The hallway nozzle was intended to simulate a crew advancing into a structure towards the burn room. Suppression from this location would be in the form of gas cooling as the stream was not aimed at the seat of the fire (indirect interior attack).  After the hallway nozzle was shutdown, there was a 15~s time period to simulate the crew advancing into the doorway of the fire room. The room nozzle was positioned just to the side of the entry door.  It was also opened and flowed for approximately 15~s. The room nozzles was aimed in the direction of the seat of the fire (direct interior attack). Table~\ref{tab:Test_Descriptions} provides an overview of the suppression tests conducted.

@@

This study focused on two key types of fire experiments; gas cooling and fire suppression. In each type of experiment a comparison was conducted between plain water and compressed air foam being discharged from nozzles of similar type with similar mass flow rates. In both types of experiments there were no significant measurable differences between the use of plain water or compressed air foam. To expand upon the 2012 and 2013 tests described in~\cite{Mitchell:1}, additional fire suppression experiments were conducted. One set of experiments used only wood based fuels, to examine the value of the reduced surface tension of the water in the CAF and its interaction with the surface of the wood fuel. A final experimental series, used a larger fire room, with more ventilation. In this final series, another change was made, the nozzle was hand held and could be moved as needed to effectively extinguish the fire. This section will discuss the overall results of the study and provide potential explanations to support them.

@@

In all of the suppression experiments, the fire was controlled with total flow of 227~L (60~gal) or less of agent. During the course of the suppression experiments the fuel packages were modified to incorporate more wood based fuels. During the planning of the experiments, some of the technical panel members felt that the synthetic fuels, such as polyurethane foam, burned mostly at the surface and therefore they were easily cooled and suppressed. Wood based fuels, however develop a char layer and therefore the burning would be ``deep seated'', which might provide a more significant suppression challenge. Since the Class A foam solution has a reduced surface tension, that could allow the solution to penetrate and cool the burning wood based fuels better, it was thought that with more exposed, burning wood surface, the better the opportunity for CAF to demonstrate a difference. As shown in Chapter~\ref{sec:fuel_load}, the wood fuel load consisted of stacks of wood pallets, plywood lined walls, and an exposed wood ``ceiling'' that was constructed of wood ``i-beams'' and oriented strand board. The experiments conducted with the wood fuel load did not demonstrate a significant difference between water and CAF.

@@

There were two different geometries used for the fire suppression experiments. Twelve of the suppression experiments were conducted in a single story structure with an entry hallway that led to a fire compartment with a floor area of 27~$m^2$ (291~$ft^2$). Within this structure, two different types of fuel loads were used. In one case, most of the fuel consisted of synthetic materials in the upholstered furniture, carpet, and padding. The other fuel load used only wood based fuels to examine the value of the reduced surface tension of the water in the CAF and its interaction with the surface of the wood fuel. In the experiments conducted in the one story building, a fixed nozzle was flowed from the hallway for approximately 15~s and then another fixed nozzle located in the fire compartment was flowed for an additional 15~s.

Seven of the suppression experiments were conducted in a two story structure.  The fire compartment was the full lower level of the structure and it had an open floor plan with an area of 51.4~$m^2$ (553~$ft^2$). At one end of the room was an open doorway, 1.8~m (6~ft) wide by
2~m (6.7~ft) high the led to a stairwell. A manual hose stream application was made from outside the doorway on the lower level. In these experiments there was a combination of wood based and synthetic fuels. The fires were allowed to transition through flashover and then the initial fire suppression attack was made. In each case the initial knockdown of the fire required 60~gal of agent or less.

@@

\begin{figure}[ht]
\includegraphics[width=.725\columnwidth]{../Figures/Script_Figures/FSE_Test_13_052013_BDP_A3}
\caption{Hallway Velocity Data from Bi-Directional Probes for Single-Story Fire Suppression Test Water~6}
\end{figure}

@@

\begin{figure}[!ht]
	\includegraphics[width=.85\columnwidth]{../Figures/Script_Figures/Test_38_West_061315_TC_A10}
	\caption{Time Series of Thermocouple Data from Array TC A10 at the Top of the Stairwell for Two Story CAF~7 Fire Suppression Test}
	\label{fig:app_caf7_tca10}
\end{figure}

>>>>

Full scale fire experiments designed to address those questions were conducted over a period of years from 2012 through 2015. The experimental arrangements and results are presented in this report. The results from both the gas cooling and the suppression of residential scale room fires showed the effectiveness of water and CAF hose streams to be similar. Under the test conditions, both agents were effective. This study did not address the previously documented attributes of CAF for improved exposure protection and improved prevention of a fire re-kindle relative to plain water.

@@

A number of fire departments throughout the US have acquired CAFS units for structural firefighting, but many have limited or stopped the use of these CAFS for structural firefighting due to operational challenges or due to lack of understanding related to fireground safety issues. This project is expected to produce a better understanding of the effectiveness and safety implications of compressed air foam (CAF) for structural firefighting through the various tasks associated with this project, including:

\begin{enumerate}
\item  Perform a literature review of available published information on the use of CAF for structural firefighting~\cite{Mitchell:1}
\item  Conduct a workshop that brings interested and experienced parties together to discuss CAF experience and to provide guidance on the proposed experimental research plan~\cite{Grant:2011} 
\item  Conduct and document well-instrumented large-scale enclosure fire tests using both CAF and plain water to evaluate the effectiveness of CAF for interior structural firefighting
\item  Conduct and document identified fire ground related experiments to evaluate differences between CAF and plain water hose lines related to issues such as effective hose stream throw and distribution, nozzle reaction force, the force to kink the hose, potential for slipping on a variety of surfaces, and flow separation~\cite{Carracino:2013,Dicus:2013,LaPolla:2012}. 
\end{enumerate}

@@

The key measures were to understand the gas cooling and fire suppression capabilities of CAF compared to water. The experimental designs were developed based on a project planning workshop, which included the discussion of incidents where firefighters had concerns or experienced minor burns while using CAF during interior attack operations in residential occupancies~\cite{Grant:2011}. Anecdotally, other fire departments or firefighters had similar incidents or impressions of CAF.

CAF generated with a solution of water and a Class A foam concentrate has been in use by wildland fire agencies since the late 1970's. The wildland fire community found that the Class A foam reduced the surface tension of water, which improved the water's ability to penetrate and soak into wildland fuels. A CAFS could turn a Class A foam solution into a blanket of bubbles that would allow the wildland firefighters to coat the fuels. The additional ``soak'' time provided by the foam layer allowed the water to penetrate the hot surface of a burning fuel and also provided additional time for water to penetrate into fuels being pre-treated in an effort to prevent ignition. An added feature of the blanket of bubbles is that firefighters can see where they had applied the agent so that they could make a more effective application. In the wildland, water for firefighting is typically a precious resource, is in limited supply, and has to be carried to where it is needed. CAF increases the effectiveness of water and may reduce the quantity required to suppress fires and prevent the ignition of adjacent fuels. Finally, the compressed air adds energy to the water stream, providing additional reach and volume to a hose stream with a relatively low mass flow rate of water. These CAF advantages combine to provide wildland firefighters with a tool that can be used to pre-treat homes and structures to protect them from an approaching fire or can be used for an exterior (indirect) fire suppression attack on a structure.

In the decades that followed the development of CAFS, a number of urban/suburban fire departments across the country began to integrate CAFS into their apparatus. For example, the Los Angeles County Fire Department incorporated CAFS into their apparatus for wildland urban interface fires and also examined using them for structural fire fighting. Other departments across the country began to experiment with CAFS for use in structural firefighting. In the few cases that were documented, the findings about the effectiveness of the CAFS were favorable. Yet, after decades of fire department trials, CAF has still not been widely incorporated into structural firefighting tactics~\cite{USFA:CAFS,Lorh:2002}. This is illustrated by the example of one community, which after putting a fleet of CAFS equipped fire engines into service, had issued an order that the CAFS should not be used after incidents occurred where firefighters were burned while performing interior suppression operations with CAF inside a structure.

There has never been a comprehensive scientific study on the use of CAF for structural fire fighting. Fire departments have conducted pilot programs to examine implementation and effectiveness of CAF; foam suppliers and equipment manufacturers have conducted minimally instrumented experiments, typically in acquired structures; and test laboratories and universities have examined small sections of the CAF puzzle, but none of these efforts has resulted in the body of knowledge needed by the fire service~\cite{Madrzykowski:3,USFA:CAFS,Colletti:1,Boyd:1,Tafreshi:1,Madrzykowski:4,Colletti:2,Tafreshi:2,Lorh:2002,Colletti:3,Kim:2012}.

Demonstration of Biodegradable, Environmentally Safe, Non-toxic Fire Suppression Liquids NISTIR 6191 was published in 1998~\cite{Madrzykowski:4}. That study examined a broad range of tests in order to determine those parameters that most critically effect fire fighting performance. The types of experiments conducted and the impact of additives relative to water are included in Table~\ref{tab:agent_tests}.

@@

Based on the results of the NIST 1998 study, the fire fighting agents, commonly referred to as Class A foams, were superior to water in reducing contact angle; cooling and penetrating wood-based fuels; mass retention on the wood fuels; and ignition inhibition of the wood fuels.  When applied in the form of foam, the water agent solutions provide enhanced radiation exposure protection to a structure relative to plain water. The foams enhanced the water by keeping the water in place on vertical surfaces such as walls and the reduced inter-facial surface tension allowed the water to penetrate absorbent fuels. These enhancements also make the agents valuable tools for mop-up. The other experiments conducted during this study---specific heat, drop size, wood crib suppression, smoke generation, and Class D fire suppression---yielded similar results for both Class A foams and plain water.

In 2005, The National Fallen Firefighters Foundation held the National Fire Service Research Agenda Symposium.  One of the needs identified was ``improved extinguishing agents and delivery systems.'' The report states that ``The primary objective of this research should be to reduce the exposure of firefighters to the risks of interior fire suppression by developing more effective extinguishing agents and systems.'' In wildland firefighting, CAFS is typically applied directly to the fuel from a remote location, typically a non-IDLH environment. In structural firefighting, crews typically have to enter the building, which contains the fire, and move under or through hot combustion products and pre-heated areas of the building (IDLH environment) as they approach the fire. In many cases, before the firefighters are able to apply the suppression agent directly to the burning fuel, they have to mitigate the hazard of the hot gas layer or try to cool interior surfaces as they move toward the fire. It is important that these distinctions and differences between exterior and interior firefighting with CAF are better understood.

Montgomery County (MD) Fire and Rescue Service (MCFRS) is one of the larger fire departments in Maryland and was a committed user of CAF with 35 CAFS front line engines. MCFRS had experienced two incidents where firefighters received burns in residential structure fires while CAF was being used to fight the fire during an interior attack. Concerns were raised within the department regarding the ability of CAF to cool fire gases and absorb heat as well as plain water. As a result, on March 4, 2010, the Fire Chief of MCFRS issued a General Order prohibiting the use of CAF during offensive interior attacks. In this same order, the Chief directed the Division Chief of Operations to continue to evaluate the use of CAF during structure fires and to report back his findings and recommendations. That order was the catalyst for this study.

@@

All of the experiments were conducted at full scale in compartments and structures that were representative of room sizes in a residential occupancy. Technical panel members whose departments routinely use CAF for structural firefighting were invited and present at many of the fire suppression experiments. The series of experiments, planned during the workshop, for examining spray density measures, gas layer cooling, and fire suppression were conducted in the fall of 2012. The data from these experiments were presented to the technical panel members. The fire suppression experiments were conducted with a fuel package that included upholstered furniture. After a review of the data, a decision was made to conduct an additional series of suppression experiments with a fuel package that was composed mainly of wood-based materials. The thinking was that the synthetic materials in the upholstered furniture burned only on the surface and were easily extinguished. wood-based fuels would char and present a deep seated fire challenge to the extinguishing agents.  After the wood-based fuel package experiments were completed in 2013, it was decided in discussions with members of the technical panel that other experimental conditions needed to be examined: room size, additional ventilation, and a manually operated handline. The issue of a larger fire compartment had also been raised during the workshop discussion, but the desire for replicate experiments and financial limitations removed the large floor plan experiments from the original plan. To accomplish the last series of experiments, additional funding was obtained from NIST, and the experiments were conducted in 2015. The fuel load contained both wood-based wall and ceiling finishes, pallets as well as carpeting and upholstered furniture.

The sections below describe the types of experiments, the experimental facilities, instrumentation, and fuel loads. Project partner MCFRS provided the CAFS equipped engines and the staffing needed to provide both the water and CAF for the experiments. These experiments were conducted at the Delaware County Emergency Services Training Center (ESTC) in Sharon Hill, PA.

\section{Description of Experiments}
\label{sec:desc_experiments}
The experimental setup for this test series is described by the three types of experiments conducted: spray density measures, gas layer cooling, and fire suppression.  Baseline operating conditions were agreed upon at the CAFS workshop for these experiments (Table~\ref{tab:op_condition})~\cite{Grant:2011}. The nozzles chosen for these experiments were Task Force Tips (TFT) Metro 1, fixed gallonage nozzles with a flow rate of 120~gpm. These are automatic nozzles which have a 7/8~inch slug-tip and can be varied from a straight stream to a wide fog stream for application. The setup also included 100~ft of 1-3/4~inch hose. Nozzles were mounted to a ``blitz-fire'' monitor device and placed inside the structures. The nozzles were controlled and operated via a gate valve from the exterior. The exterior control had a gated wye to allow the flow of CAF through a hose directed away from the test structure. When the flow of CAF was needed for the experiment, the gated wye would be switched to allow the CAF to flow into the test structure. This was done to minimize the build-up of ``dry'' foam or a slug upon the initial flow of the nozzle into the test structure.

@@

The fire suppression series included 20 experiments in the single-story experimental structures and one in the two-story structure. Twelve experiments were conducted in the single-story structures (6 in each structure) using both water and CAF application as a suppression agent. The nozzle patterns were varied and included straight stream, solid stream, and narrow fog stream. The two-story test used both water and CAF from a straight stream nozzle.

@@

The series of tests described within this report were conducted at the Delaware County ESTC in Sharon Hill, PA. A burn building and two purpose-built concrete structures are located on the grounds of the ESTC. The burn building was used for the gas cooling experiments, and the purpose-built concrete structures were used for the fire suppression experiments.

@@

The burn building, or live fire training structure, contained a two-story section and a three-story section. The building was supported with reinforced concrete beams and columns. The floors and ceilings were also concrete, and the interior walls were constructed of cement block. The walls and ceiling of the burn rooms were protected with a 25~mm (1~in) thick layer of calcium silicate insulation and was covered by 50~mm (2~in) thick concrete tile. The floor of the burn rooms was protected with fire brick. Tests were conducted in the two-story, two-room configuration, which had overall dimensions of 6.5~m (21.3~ft) by 9.2~m (30.2~ft) (cf. Fig.~\ref{fig:Delaware_County,_PA_Burn_Building_Layout}).

@@

\subsection{Concrete Block Structures}
\label{sec:Experimental Structures}
\subsubsection*{Single-Story Structures}

@@

There are different components of uncertainty in the length, mass, temperature, heat flux, gas concentration, differential pressure, gas velocity, and heat release rate reported here. Uncertainties are grouped into two categories according to the method used to estimate them. Type A uncertainties are those which are evaluated by statistical methods, and Type B are those which are evaluated by other means~\cite{Taylor&Kuyatt:1994}. Type B analysis of systematic uncertainties involves estimating the upper (+~a) and lower (-~a) limits for the quantity in question such that the probability that the value would be in the interval ($\pm$ a) is essentially 100~\%. After estimating uncertainties by either Type A or B analysis, the uncertainties are combined in quadrature to yield the combined standard uncertainty. Then, the combined standard uncertainty is multiplied by a coverage factor of two, which results in the expanded uncertainty with a 95~\% confidence interval (2$\sigma$).  For some of these components, such as the zero and calibration elements, uncertainties are derived from referenced instrument specifications. For other components, referenced research results and past experience with the instruments provided input in the uncertainty determination.

@@

There were 12 suppression tests conducted in the single story purpose-built concrete structure. Eight of the experiments used a fuel load comprised of typical household furnishings that included wooden wall paneling, two sofas, two overstuffed chairs, carpet, and padding (Section~\ref{sec:fire_suppression_furniture_fuel}). The other four experiments used wood pallet fuel load in a space with the ceiling and walls composed of wood-based materials (Section~\ref{sec:fire_suppression_pallet_fuel}). Ignition occurred in the rear of the single story structures and the available ventilation was through the front doorway and for eight of the experiments a rear window was ``opened''to provide additional ventilation (see Table~\ref{tab:Test_Descriptions}). The fires were intended to replicate post-flashover residential room and contents fires. Hose lines were pre-deployed with fixed nozzles set in two locations: one in the hallway on the approach to the burn room (hallway nozzle) and one just inside the burn room to the right side of the hallway (room nozzle). Each nozzle flowed 15~s of water or compressed air foam depending on the test. The hallway nozzle was intended to simulate a crew advancing into a structure towards the burn room. Suppression from this location would be in the form of gas cooling as the stream was not aimed at the seat of the fire (indirect interior attack).  After the hallway nozzle was shutdown, there was a 15~s time period to simulate the crew advancing into the doorway of the fire room. The room nozzle was positioned just to the side of the entry door.  It was also opened and flowed for approximately 15~s. The room nozzles was aimed in the direction of the seat of the fire (direct interior attack). Table~\ref{tab:Test_Descriptions} provides an overview of the suppression tests conducted.

@@

This study focused on two key types of fire experiments; gas cooling and fire suppression. In each type of experiment a comparison was conducted between plain water and compressed air foam being discharged from nozzles of similar type with similar mass flow rates. In both types of experiments there were no significant measurable differences between the use of plain water or compressed air foam. To expand upon the 2012 and 2013 tests described in~\cite{Mitchell:1}, additional fire suppression experiments were conducted. One set of experiments used only wood-based fuels, to examine the value of the reduced surface tension of the water in the CAF and its interaction with the surface of the wood fuel. A final experimental series, used a larger fire room, with more ventilation. In this final series, another change was made, the nozzle was hand held and could be moved as needed to effectively extinguish the fire. This section will discuss the overall results of the study and provide potential explanations to support them.

@@

In all of the suppression experiments, the fire was controlled with total flow of 227~L (60~gal) or less of agent. During the course of the suppression experiments the fuel packages were modified to incorporate more wood-based fuels. During the planning of the experiments, some of the technical panel members felt that the synthetic fuels, such as polyurethane foam, burned mostly at the surface and therefore they were easily cooled and suppressed. wood-based fuels, however develop a char layer and therefore the burning would be ``deep seated'', which might provide a more significant suppression challenge. Since the Class A foam solution has a reduced surface tension, that could allow the solution to penetrate and cool the burning wood-based fuels better, it was thought that with more exposed, burning wood surface, the better the opportunity for CAF to demonstrate a difference. As shown in Chapter~\ref{sec:fuel_load}, the wood fuel load consisted of stacks of wood pallets, plywood lined walls, and an exposed wood ``ceiling'' that was constructed of wood ``i-beams'' and oriented strand board. The experiments conducted with the wood fuel load did not demonstrate a significant difference between water and CAF.

@@

There were two different geometries used for the fire suppression experiments. Twelve of the suppression experiments were conducted in a single story structure with an entry hallway that led to a fire compartment with a floor area of 27~$m^2$ (291~$ft^2$). Within this structure, two different types of fuel loads were used. In one case, most of the fuel consisted of synthetic materials in the upholstered furniture, carpet, and padding. The other fuel load used only wood-based fuels to examine the value of the reduced surface tension of the water in the CAF and its interaction with the surface of the wood fuel. In the experiments conducted in the one story building, a fixed nozzle was flowed from the hallway for approximately 15~s and then another fixed nozzle located in the fire compartment was flowed for an additional 15~s.

Seven of the suppression experiments were conducted in a two story structure.  The fire compartment was the full lower level of the structure and it had an open floor plan with an area of 51.4~$m^2$ (553~$ft^2$). At one end of the room was an open doorway, 1.8~m (6~ft) wide by
2~m (6.7~ft) high the led to a stairwell. A manual hose stream application was made from outside the doorway on the lower level. In these experiments there was a combination of wood-based and synthetic fuels. The fires were allowed to transition through flashover and then the initial fire suppression attack was made. In each case the initial knockdown of the fire required 60~gal of agent or less.

@@

\begin{figure}[ht]
\includegraphics[width=.725\columnwidth]{../Figures/Script_Figures/FSE_Test_13_052013_BDP_A3}
\caption{Hallway Velocity Data from Bi-Directional Probes for Single-Story Fire Suppression Test Water~6}
\end{figure}
\newpage

@@

\begin{figure}[!ht]
	\includegraphics[width=.85\columnwidth]{../Figures/Script_Figures/Test_38_West_061315_TC_A10}
	\caption{Time Series of Thermocouple Data from Array TC A10 at the Top of the Stairwell for Two Story CAF~7 Fire Suppression Test}
	\label{fig:app_caf7_tca10}
\end{figure}
\clearpage
